Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces

### 2021-03-25 manuscript/manuscript.Rmd
Use better templates + handle docx output

---
title: "Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces"
short-title: "Derogations and Democratic Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: March 21, 2021
published: This is an early draft of a working paper.
git-repo: https://github.com/andrewheiss/covid-derogations
abstract: >-
  In an effort to combat the COVID-19 pandemic, most governments have imposed restrictions on civic freedoms in the interest of public health. In many cases, these emergency measures have been temporary and proportional in terms of controling the pandemic. In many other instances, however, governments have used these restrictions to suppress opposition and more permanently restrict civic space. Systematically measuring the consequences of COVID restrictions, however, is a difficult task. We examine two possible quantitative measures of the relationship between of COVID restrictions and civil society space. First, we use the Variety of Democracy project's newly released Pandemic Violations of Democratic Standards Index to explore if and how civil society restrictions predict pandemic backsliding. Second, while many countries sign international human rights treaties that ostensibly bind states to respect rights, several treaties allow for emergency derogations from these obligations. We tabulate formal human rights treaty derogations due to the pandemic and explore whether these emergency measures led to lasting declines in associational and human rights. We hope that our exploration of these two measures will provide rich data-based descriptions of the relationship between COVID restrictions and civic space that will allow for more causal work in the future.
keywords: "international human rights treaties, civil society, COVID-19, democracy"
reference-section-title: References
bibliography: bibliography.bib
csl: pandoc/csl/chicago-author-date.csl
title-page: true
editor_options: 
  chunk_output_type: console
---

@@

# Word-specific things
table_format <- ifelse(is_docx, "huxtable", "kableExtra")  # Huxtable tables
conditional_dpi <- ifelse(is_docx, 300, 300)  # Higher DPI
conditional_align <- ifelse(is_docx, "default", "center")  # Word doesn't support align

@@

library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
library(janitor)
library(ggbeeswarm)
library(patchwork)
library(sf)
library(here)

@@

pandem_clean <- readRDS(here("data", "derived_data", "pandem_clean.rds"))
pandem_single <- readRDS(here("data", "derived_data", "pandem_single.rds"))
derogations_clean <- readRDS(here("data", "derived_data", "derogations_clean.rds"))
vdem_clean <- readRDS(here("data", "derived_data", "vdem_clean.rds"))
pandem_derog <- readRDS(here("data", "derived_data", "pandem_derog.rds"))
civicus_clean <- readRDS(here("data", "derived_data", "civicus_clean.rds"))

@@

```{r tbl-emergency-responses}
pandem_single %>% 
  count(emlaw_fct_min, name = "N") %>% 
  mutate(`%` = N / sum(N)) %>% 
  mutate(`%` = percent_format(accuracy = 0.1)(`%`)) %>% 
  rename("Emergency response" = emlaw_fct_min) %>% 
  kbl(caption = "National emergency responses to the COVID-19 pandemic", 
      booktabs = TRUE, linesep = "") %>% 
  kable_styling()
```

@@

```{r tbl-derog-typology}
pandem_derog %>% 
  count(type_clean, name = "N") %>%
  rename("Type of response" = type_clean) %>% 
  kbl(caption = "Typology of possible pandemic responses across derogations and emergency responses",
      booktabs = TRUE, linesep = "") %>% 
  kable_styling()
```

@@

tbl_derog %>% 
  select(Country = country_name, `Derogation start` = derogation_start, 
         `Derogation end` = derogation_end) %>% 
  kbl(caption = "Countries that either derogated as intended or derogated with excessive human rights violations",
      booktabs = TRUE, linesep = "") %>% 
  kable_styling() %>% 
  pack_rows(index = table(fct_inorder(fct_drop(tbl_derog$type_clean_letter))))
```

@@

```{r tbl-case-summary}
tribble(
  ~Type, ~country_name, ~Summary,
  "A: Derogated as intended", "Guatemala", "Requested derogations early; derogations have been lawful and generally proportionate",
  "B: Derogated; measures not temporary, proportional, or necessary", "Armenia", "Engaged in serial derogation; has committed some extraordinary human rights abuses",
  "C: Emergency declared; no formal derogation", "Egypt", "Used pandemic as excuse to continue long-standing state of emergency and further limit civil society",
  "D: No emergency; no formal derogation", "India", "Constitutionally prohibited from declaring an emergency and unable to formally derogate; has violated treaty agreements"
) %>% 
  left_join(select(pandem_derog, country_name, pandem, panback, v2xcs_ccsi), 
            by = "country_name") %>% 
  mutate(across(c(pandem, panback, v2xcs_ccsi), ~round(., 2))) %>% 
  select(Country = country_name, Type, `Pandemic violations index` = pandem, 
         `Pandemic backsliding index` = panback, `Civil society index` = v2xcs_ccsi, Summary) %>% 
  kbl(align = c("l", "l", "c", "c", "c", "l"),
      caption = "Summary of case studies",
      booktabs = TRUE, linesep = "\\addlinespace") %>% 
  kable_styling() %>% 
  column_spec(1, width = "0.1\\\\textwidth") %>% 
  column_spec(2, width = "0.12\\\\textwidth") %>% 
  column_spec(3, width = "0.09\\\\textwidth") %>% 
  column_spec(4, width = "0.11\\\\textwidth") %>% 
  column_spec(5, width = "0.09\\\\textwidth") %>% 
  column_spec(6, width = "0.35\\\\textwidth") %>% 
  footnote(general = 
             c("Pandemic violations index ranges from 0 to 0.65; higher values represent worse violations", 
               "Pandemic backsliding index ranges from 0 to 0.5; higher values represent greater democratic backsliding",
               "Civil society index ranges from 0 to 1; higher values represent more open civil society"))
```

@@

# Appendix: Countries that did not derogate

@@

matrix_combined[,-1] %>% 
  kbl(caption = "Countries did not derogate during the pandemic",
      booktabs = TRUE, linesep = "") %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  pack_rows(index = table(matrix_combined[,1]), indent = FALSE)
```

>>>>

---
title: "Derogations and Democratic Backsliding: Exploring the Pandemic's Effects on Civic Spaces"
short-title: "Derogations and Democratic Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis & Clark College
  email: schaudhry@lclark.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: March 21, 2021
published: This is an early draft of a working paper.
code-repo: Access the code and data at <https://github.com/andrewheiss/covid-derogations>.
abstract: >-
  In an effort to combat the COVID-19 pandemic, most governments have imposed restrictions on civic freedoms in the interest of public health. In many cases, these emergency measures have been temporary and proportional in terms of controling the pandemic. In many other instances, however, governments have used these restrictions to suppress opposition and more permanently restrict civic space. Systematically measuring the consequences of COVID restrictions, however, is a difficult task. We examine two possible quantitative measures of the relationship between of COVID restrictions and civil society space. First, we use the Variety of Democracy project's newly released Pandemic Violations of Democratic Standards Index to explore if and how civil society restrictions predict pandemic backsliding. Second, while many countries sign international human rights treaties that ostensibly bind states to respect rights, several treaties allow for emergency derogations from these obligations. We tabulate formal human rights treaty derogations due to the pandemic and explore whether these emergency measures led to lasting declines in associational and human rights. We hope that our exploration of these two measures will provide rich data-based descriptions of the relationship between COVID restrictions and civic space that will allow for more causal work in the future.
keywords: "international human rights treaties, civil society, COVID-19, democracy"
reference-section-title: References
bibliography: bibliography.bib
csl: pandoc/csl/chicago-author-date.csl
title-page: true
endnotes: false
endfloat: false
mainfont: Spectral
sansfont: Noto Sans
editor_options: 
  chunk_output_type: console
---

@@

# Word-specific things
table_format <- ifelse(is_docx, "huxtable", "kableExtra")  # Huxtable tables
if (is_docx) { library(huxtable) }
conditional_dpi <- ifelse(is_docx, 300, 300)  # Higher DPI
conditional_align <- ifelse(is_docx, "default", "center")  # Word doesn't support align

@@

# Load libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
library(janitor)
library(ggbeeswarm)
library(patchwork)
library(sf)
library(here)

@@

# Load data
pandem_clean <- readRDS(here("data", "derived_data", "pandem_clean.rds"))
pandem_single <- readRDS(here("data", "derived_data", "pandem_single.rds"))
derogations_clean <- readRDS(here("data", "derived_data", "derogations_clean.rds"))
vdem_clean <- readRDS(here("data", "derived_data", "vdem_clean.rds"))
pandem_derog <- readRDS(here("data", "derived_data", "pandem_derog.rds"))
civicus_clean <- readRDS(here("data", "derived_data", "civicus_clean.rds"))

@@

```{r tbl-emergency-responses}
tbl_pandem_responses <- pandem_single %>% 
  count(emlaw_fct_min, name = "N") %>% 
  mutate(`%` = N / sum(N)) %>% 
  mutate(`%` = percent_format(accuracy = 0.1)(`%`)) %>% 
  rename("Emergency response" = emlaw_fct_min)

if (!is_docx) {
  tbl_pandem_responses %>% 
    kbl(caption = "National emergency responses to the COVID-19 pandemic", 
        booktabs = TRUE, linesep = "") %>% 
    kable_styling()
} else {
  tbl_pandem_responses %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    set_bold(1, everywhere) %>%
    set_bottom_border(1, everywhere) %>% 
    huxtable::set_caption("National emergency responses to the COVID-19 pandemic") %>% 
    huxtable::as_flextable()
}
```

@@

```{r tbl-derog-typology}
tbl_derog_typology <- pandem_derog %>% 
  count(type_clean, name = "N") %>%
  rename("Type of response" = type_clean)

if (!is_docx) {
  tbl_derog_typology %>% 
    kbl(caption = "Typology of possible pandemic responses across derogations and emergency responses",
        booktabs = TRUE, linesep = "") %>% 
    kable_styling()
} else {
  tbl_derog_typology %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    set_bold(1, everywhere) %>%
    set_bottom_border(1, everywhere) %>% 
    huxtable::set_caption("Typology of possible pandemic responses across derogations and emergency responses") %>% 
    huxtable::as_flextable()
}
```

@@

tbl_derog_a_b <- tbl_derog %>% 
  select(Country = country_name, `Derogation start` = derogation_start, 
         `Derogation end` = derogation_end)

tbl_a_b_breaks <- table(fct_inorder(fct_drop(tbl_derog$type_clean_letter)))
tbl_a_b_caption <- "Countries that either derogated as intended or derogated with excessive human rights violations"

if (!is_docx) {
  tbl_derog_a_b %>% 
    kbl(caption = tbl_a_b_caption,
        booktabs = TRUE, linesep = "") %>% 
    kable_styling() %>% 
    pack_rows(index = tbl_a_b_breaks)
} else {
  tbl_derog_a_b %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    insert_row(names(tbl_a_b_breaks[1]), "", "", after = 1) %>%
    insert_row(names(tbl_a_b_breaks[2]), "", "", 
               after = (tbl_a_b_breaks[1] + 2)) %>%
    set_bold(1, everywhere) %>%
    set_bold(2, everywhere) %>%
    set_bold((tbl_a_b_breaks[1] + 3), everywhere) %>% 
    set_bottom_border(1, everywhere) %>%
    set_bottom_border(2, everywhere) %>%
    set_bottom_border((tbl_a_b_breaks[1] + 3), everywhere) %>%
    huxtable::set_caption(tbl_a_b_caption) %>% 
    huxtable::as_flextable()
}
```

@@

```{r tbl-case-summary}
tbl_case_summary <- tribble(
  ~Type, ~country_name, ~Summary,
  "A: Derogated as intended", "Guatemala", "Requested derogations early; derogations have been lawful and generally proportionate",
  "B: Derogated; measures not temporary, proportional, or necessary", "Armenia", "Engaged in serial derogation; has committed some extraordinary human rights abuses",
  "C: Emergency declared; no formal derogation", "Egypt", "Used pandemic as excuse to continue long-standing state of emergency and further limit civil society",
  "D: No emergency; no formal derogation", "India", "Constitutionally prohibited from declaring an emergency and unable to formally derogate; has violated treaty agreements"
) %>% 
  left_join(select(pandem_derog, country_name, pandem, panback, v2xcs_ccsi), 
            by = "country_name") %>% 
  mutate(across(c(pandem, panback, v2xcs_ccsi), ~round(., 2))) %>% 
  select(Country = country_name, Type, `Pandemic violations index` = pandem, 
         `Pandemic backsliding index` = panback, `Civil society index` = v2xcs_ccsi, Summary)

tbl_case_summary_caption <- "Summary of case studies"

tbl_case_notes <- c("Pandemic violations index ranges from 0 to 0.65; higher values represent worse violations", 
                    "Pandemic backsliding index ranges from 0 to 0.5; higher values represent greater democratic backsliding",
                    "Civil society index ranges from 0 to 1; higher values represent more open civil society")

if (!is_docx) {
  tbl_case_summary %>% 
    kbl(align = c("l", "l", "c", "c", "c", "l"),
        caption = tbl_case_summary_caption,
        booktabs = TRUE, linesep = "\\addlinespace") %>% 
    kable_styling() %>% 
    column_spec(1, width = "0.1\\\\textwidth") %>% 
    column_spec(2, width = "0.12\\\\textwidth") %>% 
    column_spec(3, width = "0.09\\\\textwidth") %>% 
    column_spec(4, width = "0.11\\\\textwidth") %>% 
    column_spec(5, width = "0.09\\\\textwidth") %>% 
    column_spec(6, width = "0.35\\\\textwidth") %>% 
    footnote(general = tbl_case_notes)
} else {
  tbl_case_summary %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    set_align(everywhere, 3:5, "center") %>% 
    set_align(everywhere, c(1, 2, 6), "left") %>% 
    set_bold(1, everywhere) %>%
    set_bottom_border(1, everywhere) %>%
    huxtable::add_footnote(paste(tbl_case_notes, collapse = " • ")) %>% 
    huxtable::set_caption(tbl_case_summary_caption) %>% 
    huxtable::as_flextable()
}
```

@@

# Appendix

@@

tbl_derog_c_d <- matrix_combined[,-1]
tbl_c_d_breaks <- table(matrix_combined[,1])
tbl_c_d_caption <- "Countries that did not derogate during the pandemic"

if (!is_docx) {
  tbl_derog_c_d %>% 
    kbl(caption = tbl_c_d_caption,
        booktabs = TRUE, linesep = "") %>% 
    kable_styling(latex_options = c("hold_position")) %>% 
    pack_rows(index = tbl_c_d_breaks, indent = FALSE)
} else {
  tbl_derog_c_d %>% 
    hux() %>% 
    set_font_size(10) %>%
    set_all_padding(3) %>% 
    insert_row(names(tbl_c_d_breaks[1]), "", "", "", after = 1) %>%
    insert_row(names(tbl_c_d_breaks[2]), "", "", "",
               after = (tbl_c_d_breaks[1] + 2)) %>%
    set_bold(1, everywhere) %>%
    set_bold(2, everywhere) %>%
    set_bold((tbl_c_d_breaks[1] + 3), everywhere) %>% 
    set_bottom_border(1, everywhere) %>%
    set_bottom_border(2, everywhere) %>%
    set_bottom_border((tbl_c_d_breaks[1] + 3), everywhere) %>%
    huxtable::set_caption(tbl_c_d_caption) %>% 
    huxtable::as_flextable()
}
```
